---
title: "gordi_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gordi_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#library(gordi)
```
# Introduction
The `gordi` package was build to help poor students of community ecology deal with community data analysis, especially with ordinations and their visualization.

## General workflow
The first step to take is to calculate an ordination. The `gordi` package is compatible with the outputs of `vegan` functions, particularly:

- `rda()` for PCA and RDA
- `cca()` for CA and CCA
- `capscale()` for PCoA and db-RDA
- `metaMDS()` for NMDS
- `decorana()` for DCA

1. The `gordi_read()` function will read the resulting object of the ordination and prepare it for use by other `gordi_` functions. That said, `gordi_read()` is always the first one in the pipeline of `gordi_` functions. 

2. Decide, whether you want to display 'species' or 'sites' scores. Usually it is a good practice to duplay only one type of scores (and choose corresponding scaling in `gordi_read()`), but it is also possible to display both score types (in this case, select please `scaling = 'symmetric'`).
  
   2a. If you have decided to display species, use `gordi_species()`
   2b. If you have decided to display sites, use `gordi_sites()`


